Future challenges for Research Software Engineering

### 2025-05-14 future_challenges_for_RSEng.tex
remove 2_6 and minor fixes

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
\author{
Florian Goth\authorRef{1},
Leyla Jael Castro\authorRef{2},
Gerasimos Chourdakis\authorRef{3},
Simon Christ\authorRef{4},
Jeremy Cohen\authorRef{5},
Jean-Noël Grad\authorRef{6},
Magnus Hagdorn\authorRef{7},
Toby Hodges\authorRef{8},
Jan Linxweiler\authorRef{9},
Frank Löffler\authorRef{10},
Michele Martone\authorRef{11},
Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
Nicholas A. Del Grosso\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Würzburg-Dresden Cluster of Excellence ct.qmat, University of Würzburg, 97074, Würzburg, Germany\par
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
\autlabel{5} Imperial College London, London, UK\par
\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
\autlabel{8} The Carpentries, USA\par
\autlabel{9} Technische Universität Braunschweig, Germany\par
\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
\autlabel{12} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
}
\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all other domains.
Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

@@

\subsection{Education, and changes in the educational system}
Digital competencies in the general population continue to increase, making jumping to software engineering easier than before.

\begin{whatis}{What do we want to have achieved}{education}
School curricula have embraced the digital transformation, and have included basics of computer interaction
in every class level.
In universities, the first generations of RSE-Master graduates find their way into research departments,
and accelerate the pace of science.
\end{whatis}

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
Similar to RSEs that are known by 2035 in university boards, we expect to be the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like the HiddenRef hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have alreadyseen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernization, cleanup, and modularization,
which can pay off in faster and larger development in the future.
This is an opportunity that would not possible by short-term funded solo RSEs.

@@

\begin{story}{New career paths}{New career paths}
A story, about Ken delving into a new direction as RTP.
\end{story}

\section{What is in this future for RSEs and where to go from here?}
FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
GC-NOTE: Shall we add another story box here? Ideas what we write here?

Digitalisation will lead to changes in the way societies interact and therefore will translate
into changes in how academia perceives itself. The differentiation of job profiles in academia,
that has started with profiles such as the RSE will continue. Societies will be facing new questions, that arise
from technologies such as generally available AI systems. RSEs will very likely be doing different work,
than what they are doing today, but their core task of enhancing the scientific output of researchers, other human
beings, will stay the same.
They will still develop software artifacts together with scientists,
since Better software leads to better research,
better research leads to more better software.

>>>>

% Article frontmatter
% ===================
\title{Future challenges for Research Software Engineering} % Title of the article
%\short{} % Short title of the article (optional)
\newcommand{\authorRef}[1]{\texorpdfstring{\autref{#1}}{}}
\author{
Florian Goth\authorRef{1},
Leyla Jael Castro\authorRef{2},
Gerasimos Chourdakis\authorRef{3},
Simon Christ\authorRef{4},
Jeremy Cohen\authorRef{5},
Jean-Noël Grad\authorRef{6},
Magnus Hagdorn\authorRef{7},
Toby Hodges\authorRef{8},
Jan Linxweiler\authorRef{9},
Frank Löffler\authorRef{10},
Michele Martone\authorRef{11},
Jan Philipp Thiele\authorRef{9}\textsuperscript{,}\authorRef{12},
Nicholas A. Del Grosso\authorRef{13}
} % Authors and references to addresses
\institute{% Institutes with labels
\autlabel{1} Institut für theoretische Physik 1, University of Würzburg, 97074, Würzburg, Germany\par
\autlabel{2} ZB MED Information Centre for Life Sciences, Cologne, Germany\par
\autlabel{3} Institute for Parallel and Distributed Systems, University of Stuttgart, Stuttgart, Germany\par
\autlabel{4} Leibniz University Hannover, Department of Cell Biology and Biophysics, Computational Biology, Germany\par
\autlabel{5} Imperial College London, London, UK\par
\autlabel{6} Institute for Computational Physics, University of Stuttgart, Germany\par
\autlabel{7} Geschäftsbereich IT, Charité Universitätsmedizin Berlin, Germany\par
\autlabel{8} The Carpentries, USA\par
\autlabel{9} Technische Universität Braunschweig, Germany\par
\autlabel{10} Michael Stifel Center Jena {\&} Friedrich Schiller University Jena, Germany\par
\autlabel{11} Leibniz Supercomputing Centre, Garching, Germany\par
\autlabel{12} Weierstrass Institute, Berlin, Germany;
              Leibniz University Hannover, Institute of Applied Mathematics, Scientific Computing, Hannover, Germany\par
\autlabel{13} Institute for Experimental Epileptology and Cognition Research, Uniklinikum Bonn, Germany
}
\abstract{It's 2025 and we are, admittedly, still working on establishing and growing Research Software Engineering (RSE)
as a domain in its own right, with the hope of ensuring that it gets the recognition it deserves.
Nonetheless, as the use of digital tools becomes ever more pervasive in a researcher's day-to-day tasks, there are already people suggesting that RSE will
cease to exist as a separate field, and the specialist skills that RSEs provide to support and undertake research will
become a core part of the tooling and skill set of researchers in almost all other domains.
Are we already approaching ``the end of Research Software Engineering''?! While we believe that it is certainly too early to
already herald the demise of RSE as a separate field,
we also recognise the need to acknowledge that the future is in constant flux and
it is worthwhile discussing how RSE will/could change in response to these upcoming challenges in an open discussion.} % Abstract of the article
\keywords{Keywords go here.} % Keywords for the article

@@

\subsection{How have academic Institutions transformed by then?}
\begin{whatis}{What do we want to have achieved}{academic}
The first generations of RSE-Master graduates find their way into research departments, and accelerate the pace of science.
Similar to RSEs, that are known by 2035 in university boards, we expect the concept of RSE units to be in the minds
of academic staff at any level.
We hope to see that universities use the presence of an RSE Department actively as a means to differentiate themselves from other universities.
We hope that the central RSE units have local RSE representatives, or even small RSE groups that handle tasks within a department to oversee specific projects[FIXME: cite group paper].
We also hope, that specialized clusters form that pool the skills of certain specialised RSEs, with certain RSE units specializing in certain tasks like data management, HPC, or visualization.
At universities which have them, we want to see that researchers naturally accept them as part of their used services.
\end{whatis}
Academia, while claiming to be innovative (cite WissZeitVG here), has organisationally often adopted time-scales for change that are more reminiscient to civil services like pension funds.
Nevertheless, the presence, even today, of initiatives like \emph{the Hidden Ref} \cite{hiddenref} hints at the potential for specialisation of research tasks. This gives rise to new job descriptions within the academic system such as specialised RSEs or other RTPs.
We have already seen some specialisation trends in recent years such as
the HPC-RSE and AI-RSE which are quite recent developments, whose appeal is stimulated by large public and private investments in HPC facilities and AI technologies.
But we could imagine new RSE specialisations to emerge in the fields
of quantum computing, such as the Quantum-RSE, and in federated online platforms.
% European AI factories: https://digital-strategy.ec.europa.eu/en/policies/ai-factories
% European federation platform for HPC: https://eurohpc-ju.europa.eu/paving-way-eurohpc-federation-platform-2024-12-19_en
% European quantum computing: https://digital-strategy.ec.europa.eu/en/policies/quantum

On the other hand the existence of people with a certain specilised skillset opens up possiblities for pooling.
Universities will benefit from pooling these specialists into RSE units, and offer their services to researchers.
Hopefully, these RSE-units are endowed with enough trust by the university boards that translates into stable funding.
With longer-term funding and more trust to pick up longer term projects,
some RSEs can focus on code modernization, cleanup, and modularization,
which can pay off in faster and larger development in the future.
This is an opportunity that would not be possible by short-term funded solo RSEs.

@@

\begin{story}{New career paths}{New career paths}
A story, about Ken delving into a new direction as RTP.
\end{story}
\section{What is in this future for RSEs and where to go from here?}
%FG-NOTE: I think we can use this part to wrap things a bit up, and describe how this future then looks like.
%GC-NOTE: Shall we add another story box here? Ideas what we write here?

Digital competencies in the general population continue to increase, making jumping to software development easier than before.
But this digitalisation will also lead to changes in the way societies interact and therefore will translate into changes in how academia perceives itself.
The differentiation of job profiles in academia, that has started with profiles such as the RSE will continue.
Societies will be facing new questions, that arise from technologies such as generally available AI systems.
RSEs will very likely be doing different work,
than what they are doing today, but their core task of enhancing the scientific output of researchers -- other human
beings -- will stay the same.
They will still develop software artifacts together with scientists,
since Better software leads to better research,
better research leads to more better software.
